### 2013-12-14 mcmc.tex
toubleshooting and advice

% to-do
% -----
% - write zeroth draft
% - write first draft
% - send to friendlies for comments
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
% - post on arXiv

@@

@@

One very important note to make here,
  because it is relevant to tuning,
  is that tuning can \emph{only} take place during the burn-in phase;
  you cannot tune \emph{while} you run your final MCMC run.
Why not?
Because tuning the proposal distribution based on the past history of the chain
  violates the ``Markov'' property
  that each step depends \emph{only} on the state at the previous step.
Violating the Markov property can be very bad;
  when you violate it,
  you lose all the provable properties of MCMC
  on which all our righteous power is based.

@@

\paragraph{functional testing:}
In addition to a full set of unit tests for every part of your code
  (yeah right!\note{If you aren't unit testing, you are probably making some very big mistakes.}),
  we recommend building some end-to-end functional tests that permit you to perform MCMC
  runs with output that must meet certain expectations.
The simplest kind of functional test is to sample a known distribution,
  and check that the sampling has the moments you expect (to within tolerances).
For example, you can sample a $d$-dimensional Gaussian distribution with known non-trivial covariance,
  and check that the empirical covariance comes out as expected.
This is an easy test to fail, so a success builds confidence in any MCMC code.

@@

...Set likelihood to 1 and check that you sample your prior pdf.
...Checks properness of priors.
...Checks for think-os about parameterization.
...Make triangle plots.

>>>>

% to-do
% -----
% - where to put comments about how to structure code ln_likelihood(pars, data, I) and ln_prior(pars, I)
% - need to mention case where there is an MC integral INSIDE the likelihood function
% - need to mention testing the likelihood function
% - send to friendlies for comments
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
% - post on arXiv

@@

One note to make here
  is that you must \emph{propose}
  in the same coordinates (parameterization or transformation of parameters)
  as that in which your priors are \emph{specified},
  or else multiply in a (possibly nasty) Jacobian.
That is, if your prior is ``flat in $x$''
  but you find it easier to run your sampler by proposing steps in $\ln x$,
  if you don't modify your acceptance fraction by the Jacobian,
  your \emph{real} prior won't be flat in $x$.
These think-os can get subtle;
  the best way to avoid them
  is to write your code and your likelihood function
  and your prior pdf function and your proposal distribution
  all in precisely the same parameterization.
We will return to this point below in our comments on functional testing.

@@

One note to make here---%
  because it is relevant to tuning---%
  is that tuning can \emph{only} take place during the burn-in phase;
  you cannot tune \emph{while} you run your final MCMC run.
Why not?
Because tuning the proposal distribution based on the past history of the chain
  violates the ``Markov'' property
  that each step depends \emph{only} on the state at the previous step.
Violating the Markov property can be very bad;
  when you violate it,
  you lose all the provable properties of MCMC
  on which all our righteous power is based.

@@

\paragraph{functional testing:}
In addition to a full set of unit tests for every part of your code
  (yeah \emph{right}\note{If you aren't unit testing, you are probably making some very big mistakes.}),
  we recommend building some end-to-end functional tests that permit you to perform MCMC
  runs with output that must meet certain expectations.
The simplest kind of functional test is to sample a known distribution,
  and check that the sampling has the moments you expect (to within tolerances).
For example, you can sample a $d$-dimensional Gaussian distribution with known non-trivial covariance,
  and check that the empirical covariance comes out as expected.
This is an easy test to fail, so a success builds confidence in any MCMC code.

@@

Another important functional test is to \emph{sample the prior pdf}.
This tests the sampler,
  tests that your priors really are what you think they are
  (and there are so many ways for them \emph{not} to be\note{HOGG}),
  and tests that your prior pdfs are proper
  (functions that can be normalized).
This test is most effective as a true end-to-end test of your model
  if you structure your code such that the log probability function \code{lnprob}
  input to the MCMC scheme
  is set up as a sum of a log likelihood function \code{lnlikelihood}
  and a prior function \code{lnprior}.
The sampler can be switched to sampling the prior by a simple replacement
  of the \code{lnlikelihood} function
  with a trivial function that always returns zero.

MCMC should run well in this flat-likelihood case (priors are usually easy to sample)
  and the output sample histograms should look like what you expect
  given the prior pdfs you coded.
If the walkers run off to positive or negative infinity,
  and nothing seems to converge,
  your priors are probably not proper.
If the priors don't look as you expect,
  you either have a bug in your code,
  or else a think-o about how your priors ought to look.
Either way, diagnosis in order.

...Make triangle plots.  Great value!

Finally, if you pass the above tests but things are still incomprehensible,
  it is often a problem with the likelihood function.
Things to check include the following:
When you re-call the likelihood function with the same parameters,
  do you get \emph{exactly} the same answer?
If you don't, your likelihood function effectively depends on some random numbers;
  WE DISCUSS THIS WHERE?
When you call the likelihood function on a grid in each parameter
  (with the others fixed, say, to reasonable values),
  is the likelihood function a smooth function of the parameter,
  or at least as smooth as you expect?
Things like numerical integration,
  truncated expansions,
  or adaptive approximations
  inside likelihood functions
  can make the function noisy or jagged at small scales in parameter space.
These problems hurt optimizers more than samplers,
  but they usually point to code issues.
